### 2015-08-13 chapters/introduction.tex
expanded intro

@@

@@

@@

V4046 Sgr is a system with two stars closely orbiting each other.  Around this tight binary is a bright
protoplanetary disk.  In chapter \ref{chap:V4046Sgr} we present Submillimimeter Array (SMA) observations of this 
disk 
that resolve for the
first time a large central cavity.  In addition, we show that the distribution of gas and dust are non-commitant 
with each other: the gas disk is large and extended while the mm-sized dust grains are concentrated in a narrow 
ring.  Furthermore, we show that dust within ring is not completely cleared.  We argue that this configuration is 
the hallmark of old disk sculpted by at least one large, embedded planet.  HD163296 also has a very large, bright gaseous disk, but no resolved cavity.  In chapter \ref{chap:hd163296} we 
use 
science verification data from the Atacama Large Millimimeter Array (ALMA) to study the vertical structure of 
this disk.  We argue that resolved features in the images highlight how the temperature of the gas changes with
its distance from mid-plane and explore additional, nuanced aspects of the disk structure and how they may 
affect the observables.   In chapter \ref{chap:radial_flows} we consider what impact radial motions of gas within transition disk 
holes may have upon spectral line emission.  We discuss several indicators of such radial flows along 
with possible sources of confusion such as warps in the disk.

\section{Basics of Radio Interferometry}
\subsection{Visibilities}
\subsection{Channel Maps}
\subsection{VLBI}

@@

The Event Horizon Telescope (EHT) uses VLBI to observe supermassive black holes, obtaining resolution high enough
to resolve event horizon scales of nearby targets.  These observations provide a 
unique glimpse into one of the most extreme environments in our universe.  Using telescopes in Hawaii, Arizona, 
and California, the collaboration has already detected structure at these scales for Sgr A$^\ast$ in the galactic 
center and the giant elliptical galaxy M87 \citep{doeleman08,doeleman09}.  These data have provided indirect 
evidence for existence of an event horizon \citep{broderick15} along with constraints on the degree of 
magnetization in the jet of M87 \citep{kino15} and the orientation of Sgr A$^\ast$ \citep{broderick11}.  EHT 
observations promise to provide tests of general relativity \citep{luminet79,johannsen10,bambi13,broderick14} as 
well as new studies of black hole time variability \citep{doeleman09} and black hole accretion \citep{chan15}.

>>>>

INTRO PARAGRAPH HERE

The data presented in this thesis was obtained using radio interferometers.  This technique combines the signals
from pairs of radio telescopes to produce data that is sensitive to an angular scale inversely 
proportional to the distance between the two elements.  By assembling an array with multiple 
telescopes, an observer can accumulate information about many spatial scales and reconstruct an image.  However,
the information contained in this image will be limited to the scales originally measured by the interferometer.
Images with the best fidelity come from arrays that have elements at a wide variety of 
distances to each other and the highest resolution comes from the most widely separated telescopes.

figure: images of SMA and ALMA

Each pair of antennas in an array forms a baseline vector (measured in wavelengths) with length $D_\lambda$ and 
components $(u,v,w)$.  The signal voltages, $V_1(t)$ and $V_2(t)$, measured by the individual 
elements are correlated against each other
\begin{equation}
r(\tau) = \lim_{\tau \to \infto} \frac{1}{2T} \int_{-T}^T V_1(t) V_2^\ast(t-\tau) d\tau.
\end{equation}
This operation is often done on site with data streamed directly to a correlator and the resulting 
measurement, or visibility, is stored on disk.  However, for Very Long Baseline Interferometry (VLBI), the 
radio telescopes can be separated by entire continents and real-time correlation is not feasible.  For these
experiments, the signal voltages must be recorded and the visibilities calculated at some later time.  The Event
Horizon Telescope (EHT) a leader in the developing field of (sub)-millimeter VLBI.  Observations at these 
short wavelengths becomes more difficult as sources
tend to become fainter and the atmosphere becomes noiser, time variable, and more opaque.  Consequently, it 
has only been in the past few decades that facilities have been built to operate at these wavelengths.

The Submillimeter Array \citep[SMA][]{ho04} is one such world class array.  Commisioned in 2003, the SMA is located on the top of 
Maunakea on the Big Island of Hawaii at an altitude of 13,386 feet.
It is composed of eight 6-meter diameter dishes which, like many 
interferometers, can be moved into different configurations so that there are elements as close as 
15 meters or as far as 500 meters.  Its recievers can operate at wavelengths from 430\,$\mu$m to 2\,mm and 
historically have provided up to 4\,GHz of bandwidth.  A major upgrade to the array is increasing the bandwidth
to 8\,GHz, enabling more sensitive observations and larger spectral coverage.  To participate in VLBI 
experiments, the SMA can operate as a ``phased array'', essentially turning the interferometer into a single 
dish telescope.

Located 4450\,km away on the Chile is the Atacama Large Millimeter Array (ALMA).  This facility began early
science operations in 2011 and has been commissioned in a staged fashion where each observing cycle offers more 
capabilities.  Once completed, ALMA will host an impressive 66 antennas, 64 of will be 12\, dishes and
the remaining smaller 7\,m ones.  A fully extended configuration will include baselines up to 16\,km and the 
array will observe at wavelengths from 400\,$\mu$m to 7\,mm.  The ALMA Phasing Project (APP) will enable this
giant array to join VLBI experiments and fringes to APEX were found in a January 2015 test \citep{matthews15}.

% https://almascience.nrao.edu/documents-and-tools/alma-technical-notes-a-subset-of-documents-from-the-eoc-memo-series

@@

figure: early images of disks

@@

HERE GOES A PARAGRAPH DESCRIBING SIMPLE CONCEPTION OF DISK STRUCUTRE
figure: disk structure cartoon

@@

V4046 Sgr is a system with two stars closely orbiting each other.  Around this tight binary is a bright
protoplanetary disk.  In chapter \ref{chap:V4046Sgr} we present Submillimimeter Array (SMA) observations of this 
disk 
that resolve for the
first time a large central cavity.  In addition, we show that the distribution of gas and dust are non-commitant 
with each other: the gas disk is large and extended while the mm-sized dust grains are concentrated in a narrow 
ring.  Furthermore, we show that dust within ring is not completely cleared.  We argue that this configuration is 
the hallmark of an old disk that is being sculpted by at least one embedded planet.  HD163296 also has a 
very large, bright gaseous disk, but no resolved cavity.  In chapter \ref{chap:hd163296} we 
use 
science verification data from the Atacama Large Millimimeter Array (ALMA) to study the vertical structure of 
this disk.  We argue that resolved features in the images highlight how the temperature of the gas changes with
its distance from mid-plane and explore additional, nuanced aspects of the disk structure and how they may 
affect the observables.   In chapter \ref{chap:radial_flows} we consider what impact radial motions of gas within transition disk 
holes may have upon spectral line emission.  We discuss several indicators of such radial flows along 
with possible sources of confusion such as warps in the disk.

@@

figure: map of EHT site locations

The Event Horizon Telescope (EHT) observes supermassive black holes, obtaining resolution high enough
to resolve event horizon scales of nearby targets.  These observations provide a 
unique glimpse into one of the most extreme environments in our universe.  Using telescopes in Hawaii, Arizona, 
and California, the collaboration has already detected structure at these scales for Sgr A$^\ast$ in the galactic 
center and the giant elliptical galaxy M87 \citep{doeleman08,doeleman09}.  These data have provided indirect 
evidence for existence of an event horizon \citep{broderick15} along with constraints on the degree of 
magnetization in the jet of M87 \citep{kino15} and the orientation of Sgr A$^\ast$ \citep{broderick11}.  EHT 
observations promise to provide tests of general relativity \citep{luminet79,johannsen10,bambi13,broderick14} as 
well as new studies of black hole time variability \citep{doeleman09} and black hole accretion \citep{chan15}.
